---
title: "Reproducible Research"
author: "Nitzan Sherman"
output: html_document
---

### Loading and preprocessing the data
I loaded knitr as needed and will be using dplyr and ggplot2 for my analysis and visualizations respectively.

```{r}
library(knitr)
library(dplyr)
library(ggplot2)
opts_chunk$set(echo = TRUE,messages = FALSE)
```

I didn't do anything fancy to read the data. I used read.csv and converted data frame into tbl_df for use with the dplyr package. Before converting, I ensured that date column was changed to a date.

```{r}
raw_data <- read.csv("activity.csv")
raw_data$date <- as.character(raw_data$date)
raw_data$date <- as.Date(raw_data$date)
```

## What is mean total number of steps taken per day?

```{r}
#filters out all NA values by running complete cases on entire tbl_df.
complete_data <- raw_data %>% filter(complete.cases(.))
```

Below are the calculated total steps, mean and median per day.
```{r}
total_steps_per_day <- summarise(group_by(complete_data,date), steps = sum(steps))
```

```{r}
qplot(total_steps_per_day$steps,
      geom="histogram",
      binwidth = range(total_steps_per_day$steps)[2] / 10,
      main = "Histogram of Total Steps per Day",
      xlab = "Total Steps",
      ylab = "Count",
      fill = I("dark grey"),
      col = I("black")
      )
```

Here are the calculated mean and medians steps per day.

```{r}
mean_per_day <- summarise(group_by(complete_data,date),mean(steps))
median_per_day <- summarise(group_by(complete_data,date), median(steps))
```


## What is the average daily activity pattern?

In

```{r}
mean_per_interval <- summarise(group_by(complete_data,interval), 
                           avg = mean(steps))
```

Here is the resulting chart.
```{r}
ggplot(data=mean_per_interval, aes(x=interval, y=avg)) +
    geom_line()
```

In order to find the max value, I used the top_n function. The highest average value was at 8:35 am with about 206 steps. 

```{r}
top_n(mean_per_interval,1,avg)
```

## Imputing missing values

* Sum of Missing Values

The sum of missing values is equivalent to the difference in rows between raw data and complete data. The value is 2304.

```{r}
nrow(raw_data) - nrow(complete_data)
```

* Imputing Values Strategy

My strategy is for every NA value, to replace it with the mean steps taken for that time period. In order to do this, I combined my raw data with the mean_per_interval avg column. I used mutate to then fill out any value that was NA with the average steps.


```{r}
imputed_data <- cbind(raw_data, avg = mean_per_interval$avg)
imputed_data <- mutate(imputed_data,steps = ifelse(is.na(steps),avg,steps))
```

```{r}
imputed_total_steps_per_day <- summarise(group_by(imputed_data,date), steps = sum(steps))
imputed_mean_per_day <- summarise(group_by(imputed_data,date),mean(steps))
imputed_median_per_day <- summarise(group_by(imputed_data,date), median(steps))
```


```{r}
qplot(imputed_total_steps_per_day$steps,
      geom="histogram",
      binwidth = range(imputed_total_steps_per_day$steps)[2] / 10,
      main = "Histogram of Total Steps per Day",
      xlab = "Total Steps",
      ylab = "Count",
      fill = I("dark grey"),
      col = I("black")
      )
```

It seems like imputting the data has lessened the skew of the distribution. Looking at the totals, it seems as if the totals have gone up, while there have been very small changes to the mean and median per day.


## Are there differences in activity patterns between weekdays and weekends?

In order to do this analysis, I need to convert dates to day of the week, and then classify them as weekends or weekdays. I also need to calculate the average values per week day and weekend in order to plot them.

```{r}
imputed_data <- mutate(imputed_data,day = weekdays(date))
imputed_data <- mutate(imputed_data,day_type = ifelse(day %in% c("Saturday","Sunday"),0,1))
imputed_data$day_type <- factor(imputed_data$day_type,labels = c("weekend","weekday"))
mean_per_interval_per_day_type <- summarise(group_by(imputed_data,day_type,interval),avg = mean(steps))
```

Here is the plot comparing weekend activity to weekday activity.
```{r}
qplot(x = interval,
      y = avg,
      data = mean_per_interval_per_day_type,
      xlab = "Interval",
      ylab = "Steps",
      main = "Average Activity Steps per Interval",
      facets = day_type ~.,
      geom = c("line","smooth")
      )
```

There's a few things that stick out. Notice how weeekend activity tends to be more active in the afternoon, while early morning is where the most activity is for weekdays. This intuitively makes sense as mornings are when people are rusing out the door. On the weekends, the afternoon is when people socialize.